---
title: "earthquakeGeoms"
author: "Rony Sulca"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The package `earthquakeGeoms` has been built to clean and display the earthquake data from NOAA (The National Centers for Environmental
Information). The dataset can be found here: https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1

The Package is divided into 3 sections:

* Cleaning the Dataset
* Displaying the Earthquakes through the geom_timeline and geom_timeline_label functions.
* Displaying the Earthquakes by their geographical location, through leaflet.

These 3 sections give you a comprehensive way to analyze and visualize the Earthquake data. 

## Cleaning the Dataset

The cleaning part of the package is composed 2 functions:

### eq_location clean

`eq_location_clean` takes an argument `data`, and returns that same data with its column LOCATION_NAME cleaned.
The cleaning is done by splitting the character string by the `:` symbol. Then, the latest of the strings is selected (
the rigth-most string). Finally, the strings are titled, that is, only the initial letter is uppercase while the rest
remain lowercase (for each word).

```{r, echo = FALSE}
if (require(earthquakeGeoms)){
  library(earthquakeGeoms)
}
```

```{r, warning=FALSE}
test_dataframe = data.frame(order = 1:10,
                            LOCATION_NAME = c("Peru: Lima", "Mexico: Tijuana",
                                              "CUBA: HABANA", "SRI LANKA: IslanD",
                                              "EU: SPain: unknown", "US: Maine",
                                              "US: California: Los Angeles", " Ecuador: Bogota",
                                              "Ivory COAST: ISLAND 3", "China: HONG KONG"),
                            stringsAsFactors = F)

test_dataframe = eq_location_clean(test_dataframe)

test_dataframe
```

### eq_clean_data

`eq_clean_data` takes an argument `data`, same as `eq_location_clean`, and returns a cleaned dataset with the `eq_location_clean`
applied to it. Furthermore, the LATITUDE, LONGITUDE, DEATHS, TOTAL_DEATHS columns are converted to numeric type. The data points whose
YEAR columns is below 0 (BC years) are removed. The DAY, and MONTH columns have their NA values converted to 1. Finally, a new DATE column is created by combining the DAY, MONTH, and YEAR columns and
converting them to Date type.


```{r}
test_dataframe = data.frame(LOCATION_NAME = c("Peru: Lima", "Mexico: Tijuana",
                                              "CUBA: HABANA", "SRI LANKA: IslanD",
                                              "EU: SPain: unknown", "US: Maine",
                                              "US: California: Los Angeles"),
                            YEAR = c(2150,-2000,2000,1999,-1610,1566,1450),
                            MONTH = c(NA,3,NA,1,9,6,5),
                            DAY = c(NA,NA,1,4,NA,2,NA),
                            LATITUDE = c(31.1,38,35.683,34.1,36.4,31.5,35.5),
                            LONGITUDE = c(35.5,58.2,35.8,43.2,25.4,35.3,25.5),
                            DEATHS = c(NA,1,2,8,33,NA,NA),
                            TOTAL_DEATHS = c(NA,1,NA,9,NA,25,NA),
                            stringsAsFactors = F)

test_dataframe = eq_clean_data(test_dataframe) # Calling the function

test_dataframe
```


## GeomTimeline & GeomTimelineLabel

The newly created geoms, Geom Timeline & Geom Timeline Label, display the earthquake data through points of varying color, size, and location (based on coordinates).

### GeomTimeline

`GeomTimeline` is a new Geom made of pointGrobs and lineGrobs, which display the input `data` as points. These points vary based on the colour and size arguments. If a y-axis is given, the points will be displayed in different y-coordinates. Furthermore, a straight line is drawn from the minimum x-value to the maximum x-value. If there are y-coordinates, the line is drawn for each y. This Geom works once it is called by its layer function `geom_timeline`.

### geom_timeline

`geom_timeline` is the layer function for the GeomTimeline geom. To call the Geom, the `data` argument must be supplied.
Furthermore, the colour and size arguments will indicate the color and size of the data points. If, the y argument is supplied,
the data points points will be grouped into n(y) groups, each displaying its own line of points.


```{r, fig.width=7.2, fig.height=3}
test_dataframe = data.frame(LOCATION_NAME = c("Peru: Lima", "Mexico: Tijuana",
                                              "CUBA: HABANA", "SRI LANKA: IslanD",
                                              "EU: SPain: unknown", "US: Maine",
                                              "US: California: Los Angeles"),
                            YEAR = c(2150,-2000,2000,1999,-1610,1566,1450),
                            MONTH = c(NA,3,NA,1,9,6,5),
                            DAY = c(NA,NA,1,4,NA,2,NA),
                            LATITUDE = c(31.1,38,35.683,34.1,36.4,31.5,35.5),
                            LONGITUDE = c(35.5,58.2,35.8,43.2,25.4,35.3,25.5),
                            DEATHS = c(NA,1,2,8,33,NA,NA),
                            TOTAL_DEATHS = c(NA,1,NA,9,NA,25,NA),
                            EQ_PRIMARY = c(3.6, 4.4,6.5, 2.7, 5.5, 6.3, 7.4),
                            stringsAsFactors = F)
test_dataframe = eq_clean_data(test_dataframe)

ggplot2::ggplot(test_dataframe, ggplot2::aes(DATE,
                                      colour = TOTAL_DEATHS,
                                      size = as.numeric(EQ_PRIMARY) )) +
  geom_timeline() +
  ggplot2::guides(size = ggplot2::guide_legend(title = "Ritcher scale value")) +
  ggplot2::scale_colour_continuous(name = "# of DEATHS") +
  ggplot2::theme_classic()

```


### GeomTimelineLabel

`GeomTimelineLabel` is a new geom made of lineGrobs and textGrobs, which label the points from the `data` argument. The location of these labels are based on the provided `x` argument. The text used to label should be given as the `label` argument. The textGrob is positioned above the x-coordinate of the points, right above a vertical line drawn at the point. If y-coordinates are given, the vertical lines drawn will be smaller in size, depending on how many unique y-coordinates there are.

### geom_timeline_label

`geom_timeline_label` is the layer function to call the GeomTimelineLabel geom. This label works perfectly with the previously defined `geom_timeline` function, and inherits its parameters as with any other ggplot. The extra argument needed is the `label` argument. Another optional argument used is `nmax`, which describes how many top labels to use in the annotation. By 'top' label, we ferer to the data points with the higher `nmax` values based on their `size` attribute. Note: If you need to provide a size, nmax, or other argument as a single value, rather than a column of the data, provide it outside the aes function.


```{r, fig.width=7.2, fig.height=3}
ggplot2::ggplot(test_dataframe, ggplot2::aes(DATE,
                                      colour = TOTAL_DEATHS,
                                      size = as.numeric(EQ_PRIMARY) )) +
  geom_timeline() +
  geom_timeline_label(ggplot2::aes(label = test_dataframe$LOCATION_NAME)) + # Labeling
  ggplot2::guides(size = ggplot2::guide_legend(title = "Ritcher scale value")) +
  ggplot2::scale_colour_continuous(name = "# of DEATHS") +
  ggplot2::theme_classic()

```


## Leaflet Functions

The following 2 functions were built to display the earthquake data by their geographical location through a leaflet map.

### eq_map

`eq_map` is a function that displays the data points in the input `data` argument, based on its LATITUDE and LONGITUDE columns. The data points will be displayed as circles in the map. The second argument, `annot_col` decides which column in the dataset will be used as the pop-up annotation of the leaflet map.

```{r, fig.width=7.2, fig.height=3}

## Data Setup

test_dataframe = data.frame(LOCATION_NAME = c("Peru: Lima", "Mexico: Tijuana",
                                              "CUBA: HABANA", "SRI LANKA: IslanD",
                                              "EU: SPain: unknown", "US: Maine",
                                              "US: California: Los Angeles"),
                            YEAR = c(2150,-2000,2000,1999,-1610,1566,1450),
                            MONTH = c(NA,3,NA,1,9,6,5),
                            DAY = c(NA,NA,1,4,NA,2,NA),
                            LATITUDE = c(31.1,38,35.683,34.1,36.4,31.5,35.5),
                            LONGITUDE = c(35.5,58.2,35.8,43.2,25.4,35.3,25.5),
                            DEATHS = c(NA,1,2,8,33,NA,NA),
                            TOTAL_DEATHS = c(NA,1,NA,9,NA,25,NA),
                            EQ_PRIMARY = c("3.6", "4.4","6.5", "2.7", "5.5", "6.3", "7.4"),
                            stringsAsFactors = F)
test_dataframe = eq_clean_data(test_dataframe)

## Function Call

m = eq_clean_data(test_dataframe)

eq_map(m, annot_col = "DATE")

```

### eq_create_label

`eq_create_label` creates a new column called 'popup_text' from the LOCATION_NAME, EQ_PRIMARY, and TOTAL_DEATHS columns of the input argument `data`. The resulting column is a character vector formatted in HTML-format, where the column LOCATION_NAME is titled 'Location', the column 'EQ_PRIMARY' is titled 'Magnitude', and the column 'TOTAL_DEATHS' is titled 'Total deaths'. If any of the dataset's rows is missing either of the column values, then that particular annotation is skipped in the html string.


```{r, fig.width=7.2, fig.height=3}

## Function Call

m$popup_text = eq_create_label(m) # Creating the Labels

eq_map(m, annot_col = "popup_text")

```




